Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2015-04-01 paint-contour-tree.tex
small changes

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$. 
For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e. a contour.

@@

Suppose we have $f: \MM \mapsto \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang et al build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ from Chiang et al~\cite{cllr-sooscctmp-05}, based on a construction of Bajaj et al~\cite{BaKr+98}.

But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for some instances
and characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald et al~\cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour trees definitions and algorithms.

>>>>

Geometric data (potentially massive is size) is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$. 
For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e. a contour.

@@

Suppose we have $f: \MM \mapsto \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can one beat this sorting bound for some instances
and characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour trees definitions and algorithms.
